\vspace{2cm} \Large \bf Seasonal influenza circulation patterns and future projections

### 2016-02-17 auspice/reports/tex/feb-2016.tex
Fix typo.
Include notes on HA1 numbering of HA2 substitutions.
Add numbering notes and adjust frequency estimation parameters.
Update Feb report TeX.

\date{February 15, 2016}

@@

\textbf{We expect clade 3c2.a viruses to continue to predominate over 3c3.a and 3c3.b viruses. Diversity within 3c2.a is beginning to accrue with the emergence of several clades of appreciable frequency, the most notable of which is the 171K/77V clade that has very recently risen to high frequency.}

@@

Within clade 3c2.a, there have emerged multiple circulating amino acid variants including HA1:114T, HA1:142K/197R and (HA1:171K, HA2:77V/155E) (Fig.\ \ref{H3N2_3c2a_tree}). The substitution 171K has emerged twice within 3c3.a. The earlier substitution did not spread far, however, the recent substitution has occurred alongside HA2:77V/155E and is present in more recent viral isolates.

@@

These variants show varying dynamics (Fig.\ \ref{H3N2_mutations}). The 114T variant peaked in frequency around July 2015 and similarly the 142K variant has been around for more than a year with a low but even geographic distribution. Consequently, we expect neither of these variants to be under strong positive selection. The T160K substitution is observed scattered within clade 3c2.a and overrepresented in sequences obtained from cultured virus. However, we observe a recent rapid increase in frequency of (HA1:171K, HA2:77V/155E) especially in Asia and North America. This has been the fastest growing variant within 3c2.a that we have observed. This clade also shows rapid phylogenetic branching \cite{neher2014predicting}, corroborating its rapid rise. Barring substantial changes in other clades, we predict the (HA1:171K, HA2:77V/155E) variant to dominate.

@@

Due to poorly performing HI assays on 3c2.a samples and the lack of recent data, antigenic diversity within 3c2.a is hard to assess. The data available to us do not suggest significant antigenic evolution. Here we show recent antigenic data using sera against A/HongKong/5738/2014 from the Crick Worldwide Influenza Centre alongside model fits \cite{neher2015prediction} (Fig.\ \ref{H3N2_HI}). Cooler color indicate greater antigenic similarity. Note that we lack HI data for (HA1:171K, HA2:77V/155E) viruses.

@@

Within clade 6b, genetic variants have begun to emerge, most notably HA1:84N/162N/216T, referred to here as clade 6b.1, and (HA1:152T, HA2:174E), referred to here as clade 6b.2 (Fig.\ \ref{H1N1pdm_tree}).

@@

Within clade 3, HA1:172Q has predominated during the whole of 2015, while the subclade HA1:172Q/251V has been slowly rising to frequencies close to 100\% in all geographic regions (Fig.\ \ref{Yam_mutations}). The HI model \cite{neher2015prediction} assigns a modest titer drop of about 0.3 units to the M251V branch. Within the clade 251V there are multiple subclades of intermediate size, one of which predominates in China. None of these exhibit a particularly striking dynamics.

>>>>

\date{February 16, 2016}

@@

\textbf{We expect clade 3c2.a viruses to continue to predominate over 3c3.a and 3c3.b viruses. Diversity within 3c2.a is beginning to accrue with the emergence of several clades of appreciable frequency, the most notable of which is the new 171K clade that has very recently risen to high frequency.}

@@

Within clade 3c2.a, there have emerged multiple circulating amino acid variants including HA1:114T, HA1:142K/197R and (HA1:171K, HA2:77V/155E) (Fig.\ \ref{H3N2_3c2a_tree}). This last clade corresponds to HA1:171K/406V/484E. The substitution 171K has emerged twice within 3c3.a. The earlier substitution did not spread far, however, the recent substitution has occurred alongside HA2:77V/155E and is present in more recent viral isolates. Within clade 142K/197R there is a subclade of significant size bearing 168V.

@@

These variants show varying dynamics (Fig.\ \ref{H3N2_mutations}). The 114T variant peaked in frequency around July 2015 and similarly the 142K variant has been around for more than a year with a low but even geographic distribution. Consequently, we expect neither of these variants to be under strong positive selection. The T160K substitution is observed scattered within clade 3c2.a and overrepresented in sequences obtained from cultured virus.

We observe a recent rapid increase in frequency of (HA1:171K, HA2:77V/155E) especially in Asia and North America. This has been the fastest growing variant within 3c2.a that we have observed. This clade also shows rapid phylogenetic branching \cite{neher2014predicting}, corroborating its rapid rise. Additionally, the subclade bearing HA1:142K/197R/168V has recently risen in frequency, although not as rapidly as the (HA1:171K, HA2:77V/155E) clade. We estimate clade HA1:142K/197R/168V has gone from 1\% global frequency in March 2015 to 19\% global frequency today, while clade (HA1:171K, HA2:77V/155E) has gone from 1\% global frequency in June 2015 to 29\% global frequency today.

Barring substantial changes in other clades, we predict the (HA1:171K, HA2:77V/155E) variant to dominate.

@@

Due to poorly performing HI assays on 3c2.a samples and the lack of recent data, antigenic diversity within 3c2.a is hard to assess. The data available to us do not suggest significant antigenic evolution. Here we show recent antigenic data using sera against A/HongKong/5738/2014 from the Crick Worldwide Influenza Centre alongside model fits \cite{neher2015prediction} (Fig.\ \ref{H3N2_HI}). Cooler color indicate greater antigenic similarity. Note that we lack HI data for (HA1:171K, HA2:77V/155E) viruses. Epitope mutation count would predict that HA1:142K/197R/168V is more drifted than (HA1:171K, HA2:77V/155E) with 2 epitope mutations vs 1 epitope mutation, although this requires confirmation through HI or other antigenic characterization.

@@

Within clade 6b, genetic variants have begun to emerge, most notably HA1:84N/162N/216T, referred to here as clade 6b.1, and (HA1:152T, HA2:174E), referred to here as clade 6b.2 (Fig.\ \ref{H1N1pdm_tree}). The 6b.2 clade corresponds to HA1:152T/501E.

@@

Within clade 3, HA1:172Q has predominated during the whole of 2015, while the subclade HA1:172Q/251V has been slowly rising to frequencies close to 100\% in all geographic regions (Fig.\ \ref{Yam_mutations}). The HI model \cite{neher2015prediction} assigns a modest titer drop of about 0.3 units to the M251V branch. Within the clade 251V there are multiple subclades of intermediate size, one of which predominates in China. None of these exhibit particularly striking dynamics.
